---
title: LSI Section 1.1
output:
  html_document:
    code_folding: hide
header-includes:
  - \usepackage{color}
---

```{r, include=FALSE}
require(tgsify)
```

\[
\mu \sim N(\theta, A) \\
z|\mu \sim N(\mu, B)
\]

\[
\begin{align*}
f_{\mu|z} \propto &\ f_{\mu}f_{z|\mu} \\
\log f_{\mu|z} =&\ \text{[stuff]} + \left\{ - \frac{1}{2} (\mu - \theta)^t A^{-1}(\mu-\theta)\right\} + \left\{-\frac{1}{2}(z-\mu)^tB^{-1}(z-\mu) \right\} \\
&\ \text{Note, we only care about terms involving } \mu\\
&\ \text{[stuff]} + \left\{ - \frac{1}{2} ({\color{red}\mu} - \theta)^t A^{-1}({\color{red}\mu}-\theta)\right\} + \left\{-\frac{1}{2}(z-{\color{red}\mu})^tB^{-1}(z-{\color{red}\mu}) \right\} \\
&\ \text{[more stuff]} -\frac{1}{2}\mu^tA^{-1}\mu + \mu^tA^{-1}\theta + \mu^tB^{-1}z - \frac{1}{2}\mu^tB^{-1}\mu\\
&\text{Group quadratic and linear terms}\\
& \text{[stuff]} -\frac{1}{2}\mu^t\left[A^{-1} + B^{-1}\right]\mu + \mu^t\left[A^{-1}\theta + B^{-1}z\right]\\
&\text{Multiply by the identity}\\
& \text{[stuff]} -\frac{1}{2}\mu^t\left[A^{-1} + B^{-1}\right]\mu + \mu^t\left[A^{-1} + B^{-1}\right]\underbrace{\left[A^{-1} + B^{-1}\right]^{-1}\left[A^{-1}\theta + B^{-1}z\right]}_{q}\\
& \text{[stuff]} -\frac{1}{2}\mu^t\left[A^{-1} + B^{-1}\right]\mu + \mu^t\left[A^{-1} + B^{-1}\right]q\\
& \text{[even more stuff not invovling}\ \mu] -\frac{1}{2}(\mu - q)^t\left[A^{-1} + B^{-1}\right](\mu - q)\\
\mu|z \sim & N(\left[A^{-1} + B^{-1}\right]^{-1}\left[A^{-1}\theta + B^{-1}z\right], [A^{-1} + B^{-1}]^{-1})
\end{align*}
\]

When $A = aI$ and $B = bI$, then
\[
\mu|z \sim N\left(\frac{ab}{a+b}\left[\frac{1}{a}\theta + \frac{1}{b}z\right],\ \frac{ab}{a + b} \cdot I \right)
\]

If, as in the book, $\theta = 0$ and $B = I$, then
\[
\mu|z \sim N\left(\frac{a}{a+1}\cdot z,\ \frac{a}{a + 1} \cdot I \right)
\]


## Posterior when θ is e, a is π, and b is the golden ratio
```{r}
set.seed(2345)
theta <- exp(1)
a <- pi
b <- (1 + sqrt(5))/2 #phi


mu <- rnorm(1, theta, sqrt(a))
z <- rnorm(1, mu, sqrt(b))

mean <- a*b/(a+b)*(theta/a + z/b)
sigma <- a*b/(a+b) %>% sqrt

plotstyle(style = upright)
par(mar = c(2,2,2,1))
curve(
    dnorm(x, mean, sigma)
  , from = mean - 3*sigma
  , to = mean + 3*sigma
  , ylab = ""
  , xlab = ""
  , lwd = 3
)

abline(v = c(z, theta), lwd = 2, col = "darkblue")
axis(3, at = c(z, theta), label = c("z", "θ"))
abline(v = mean, col = "red", lty = 2)
axis(3, at = mean, label = "E(μ|z)", line = 1, tcl = 0)
```