---
title: LSI Section 1.1
output:
  html_document:
    code_folding: hide
header-includes:
  - \usepackage{color}
---

```{r, include=FALSE}
require(tgsify)
```

## From equation 1.1 to 1.10 (with slightly different notation)

\[
\mu \sim N(\theta, A) \\
z|\mu \sim N(\mu, B)
\]

\[
\begin{align*}
f_{\mu|z} \propto &\ f_{\mu}f_{z|\mu} \\
\log f_{\mu|z} =&\ \text{[stuff]} + \left\{ - \frac{1}{2} (\mu - \theta)^t A^{-1}(\mu-\theta)\right\} + \left\{-\frac{1}{2}(z-\mu)^tB^{-1}(z-\mu) \right\} \\
&\ \text{Note, we only care about terms involving } \mu\\
&\ \text{[stuff]} + \left\{ - \frac{1}{2} ({\color{red}\mu} - \theta)^t A^{-1}({\color{red}\mu}-\theta)\right\} + \left\{-\frac{1}{2}(z-{\color{red}\mu})^tB^{-1}(z-{\color{red}\mu}) \right\} \\
&\ \text{[more stuff]} -\frac{1}{2}\mu^tA^{-1}\mu + \mu^tA^{-1}\theta + \mu^tB^{-1}z - \frac{1}{2}\mu^tB^{-1}\mu\\
&\text{Group quadratic and linear terms}\\
& \text{[stuff]} -\frac{1}{2}\mu^t\left[A^{-1} + B^{-1}\right]\mu + \mu^t\left[A^{-1}\theta + B^{-1}z\right]\\
&\text{Multiply by the identity}\\
& \text{[stuff]} -\frac{1}{2}\mu^t\left[A^{-1} + B^{-1}\right]\mu + \mu^t\left[A^{-1} + B^{-1}\right]\underbrace{\left[A^{-1} + B^{-1}\right]^{-1}\left[A^{-1}\theta + B^{-1}z\right]}_{q}\\
& \text{[stuff]} -\frac{1}{2}\mu^t\left[A^{-1} + B^{-1}\right]\mu + \mu^t\left[A^{-1} + B^{-1}\right]q\\
& \text{[even more stuff not invovling}\ \mu] -\frac{1}{2}(\mu - q)^t\left[A^{-1} + B^{-1}\right](\mu - q)\\
\mu|z \sim & N(\left[A^{-1} + B^{-1}\right]^{-1}\left[A^{-1}\theta + B^{-1}z\right], [A^{-1} + B^{-1}]^{-1})
\end{align*}
\]

When $A = aI$ and $B = bI$, then
\[
\mu|z \sim N\left(\frac{ab}{a+b}\left[\frac{1}{a}\theta + \frac{1}{b}z\right],\ \frac{ab}{a + b} \cdot I \right)
\]

If, as in the book, $\theta = 0$ and $B = I$, then
\[
\mu|z \sim N\left(\frac{a}{a+1}\cdot z,\ \frac{a}{a + 1} \cdot I \right)
\]


### Example: Posterior when θ is e, a is π, and b is the golden ratio
```{r}
set.seed(2345)
theta <- exp(1)
a <- pi
b <- (1 + sqrt(5))/2 #phi


mu <- rnorm(1, theta, sqrt(a))
z <- rnorm(1, mu, sqrt(b))

mean <- a*b/(a+b)*(theta/a + z/b)
sigma <- a*b/(a+b) %>% sqrt

plotstyle(style = upright)
par(mar = c(2,2,2,1))
curve(
    dnorm(x, mean, sigma)
  , from = mean - 3*sigma
  , to = mean + 3*sigma
  , ylab = ""
  , xlab = ""
  , lwd = 3
)

abline(v = c(z, theta), lwd = 2, col = "darkblue")
axis(3, at = c(z, theta), label = c("z", "θ"))
abline(v = mean, col = "red", lty = 2)
axis(3, at = mean, label = "E(μ|z)", line = 1, tcl = 0)
```

## Equation 1.15

Let $\tau(z)$ be a generic estimator of $\mu$. Let $\hat{\mu}$ denote the MLE estimate of $\mu$.

Let $R\left(\mu, \tau(z)\right)$ denote risk (or frequentist risk), i.e.,
\[
\begin{align*}
R\left(\mu, \tau(z)\right) =\ &E_{z|\mu}\left[L\left(\mu, \tau(z) \right) \right] \\
& \int L\left(\mu, \tau(z) \right) p(z|\mu)\, dx.
\end{align*}
\]

Bayes risk, denote with $B$, is the frequentist risk averaged over the prior.

\[
\begin{align*}
B(\tau(z)) =\ &E_{\mu}\left[E_{z|\mu}\left[L\left(\tau(z), \mu\right) \right] \right] \\
&\int \int L\left(\tau(z), \mu\right) p(z|\mu)\,dx\,p(\mu)\,d\mu
\end{align*}
\]

In the case of the MLE, $\hat{\mu} = z$, the frequentist risk with squared error loss is
\[
\begin{align*}
R\left(\mu, \hat{\mu}\right) =\ & E_{z|\mu}\left[ (z - \mu)^t(z-\mu) \right]\\
& \text{Note if } B = I \text{ then } z-\mu \sim N(0, I) \text{ which implies } (z - \mu)^t(z-\mu) \sim \chi^2_N \\
=\ & N
\end{align*}
\]

Bayes risk of the MLE, then, is
\[
B(\hat{\mu}) = E_\mu[N] = N.
\]

## Equation 1.18

The posterior mean estimator is (when $B = I$ and $A = aI$)
\[
\dot\mu = E[\mu|z] = \frac{a}{a+1}\left[\frac{1}{a}\theta + z\right].
\] 

The frequentist risk of $\dot\mu$ is
\[
\begin{align*}
R(\mu, \dot\mu) =\ & E_{z|\mu}\left[ (\dot\mu - \mu)^t(\dot\mu - \mu) \right]\\
&\text{Note } \dot\mu - \mu = \frac{a}{a+1}\left[\frac{1}{a}\theta + z\right] - \mu \sim N\left(\frac{1}{a+1} \cdot (\theta -\mu), \frac{a^2}{(a+1)^2}\cdot I \right)\\
&\text{So, } \frac{(a+1)^2}{a^2}(\dot\mu - \mu)^t(\dot\mu - \mu) \sim \chi^2_{N,\lambda} \text{ where } \lambda = \frac{1}{2a^2} (\theta - \mu)^t(\theta - \mu)\\
=\ & \frac{a^2}{(a+1)^2}E_{z|\mu}\left[ \frac{(a+1)^2}{a^2}(\dot\mu - \mu)^t(\dot\mu - \mu) \right] \\
=\ & \frac{a^2}{(a+1)^2}\left(N + 2\lambda\right) \\
=\ & \frac{a^2}{(a+1)^2}\left(N + \frac{1}{a^2} (\theta - \mu)^t(\theta - \mu)\right) \\
=\ & \frac{a^2}{(a+1)^2} N + \frac{1}{(a+1)^2} (\theta - \mu)^t(\theta - \mu) \\
\end{align*}
\]

The Bayesian risk of $\dot\mu$ is
\[
\begin{align*}
B(\dot\mu) =\ &E_{\mu}\left[\frac{a^2}{(a+1)^2} N + \frac{1}{(a+1)^2} (\theta - \mu)^t(\theta - \mu) \right] \\ 
&\text{Note }\frac{1}{\sqrt{a}}(\theta - \mu) \sim N(0,I) \text{ so }  \frac{1}{a} (\theta - \mu)^t(\theta - \mu) \sim \chi^2_N\\
=\ & \frac{a^2}{(a+1)^2} N  + \frac{a}{(a+1)^2} E_{\mu}\left[ \frac{1}{a} (\theta - \mu)^t(\theta - \mu) \right] \\
=\ & \frac{a^2}{(a+1)^2} N + \frac{a}{(a+1)^2} N \\
=\ & \frac{a}{(a+1)} N
\end{align*}
\]
