---
title: LSI Section 5 vfw
output:
  html_document:
    toc: TRUE
    toc_float:
      collapsed: yes
header_includes:
  - \usepackage{amsmath, mathtools}
  - \usepackage[mathscr]{euscript}
---
   
# Section 5.1

```{r}
knitr::opts_knit$set(results = 'markup')
```

## Estimate $\pi_0$ using section 4.5 methods

### Prostate data

The `prostz.RData` consists of 6,033 z-values from the prostate cancer data analysis example. 

```{r}
load(file.path("/Users/valeriewelty/Dropbox/Vandy/SEDS Lab Group/lsi", "prostz.RData"))
length(prostz)
prostz = prostz[which(abs(prostz) <= 4.5)]
```

In this example, we take $\mathscr{A}_0$ to be the central $\alpha_0$ proportion of the $f_0$ distribution as on page 60. Here we use $\alpha_0 = 0.5$ and $f_0$ is $\mathscr{N}(0, 1)$ such that   

\begin{align}
\mathscr{A}_0 & = \left[ \Phi^{-1} (0.5 - \alpha_0/2), \Phi^{-1}(0.5+\alpha_0/2) \right] \\
& = \left[ \Phi^{-1} (0.5 - 0.5/2), \Phi^{-1}(0.5+0.5/2) \right] \\
& = \left[ \Phi^{-1} (0.25), \Phi^{-1}(0.75) \right] \\
& = \left[ -0.6745, 0.6745 \right] 
\end{align}

```{r}
(a0 = 0.5)
(A0 = qnorm(0.5 + c(-1, 1) * a0/2))
```

We "believe" that $f_1(z)$ is zero for this subset of the sample space, i.e. $f_1(z) = 0$ for $z \in \mathscr{A} = [ -0.6745, 0.6745 ]$. By this assumption, all non-null cases have $z$-values outside of $\mathscr{A}_0$ (called the _zero assumption_). Recall  $N_+(\mathscr{A}_0)$ is the observed number of $z_i$ values in $\mathscr{A}_0$, and here is equal to 2,790. 

```{r}
length(which(prostz >= A0[1] & prostz <= A0[2]))
```
 
Recall that $E[N_+(\mathscr{A}_0)] = \pi_0 \cdot N \cdot F_0(\mathscr{A}_0)$ (4.45), suggesting the estimator 

\begin{align}
\widehat{\pi}_0 & = N_+(\mathscr{A}_0)/(N \cdot F_0(\mathscr{A}_0)) \\
& = \frac{2,790}{6,033 \cdot F_0(\mathscr{A}_0)}
\end{align}

```{r}
N = length(prostz)
N_plus_A0 = length(which(prostz >= A0[1] & prostz <= A0[2]))

(pi0 = N_plus_A0/(N * (pnorm(A0[2]) - pnorm(A0[1]))))
```

### DTI data

For the DTI data set, $N = 15,443$ and $N_+(\mathscr{A}_0) = 7,165$ and therefore $\widehat{\pi}_0 = $. 

```{r}
load(file.path("/Users/valeriewelty/Dropbox/Vandy/SEDS Lab Group/lsi", "DTIdata.Rda"))
DTIz = DTIdata[, 'z.value']
  
(N = length(DTIz))
(N_plus_A0 = length(which(DTIz >= A0[1] & DTIz <= A0[2])))
(pi0 = N_plus_A0/(N * (pnorm(A0[2]) - pnorm(A0[1]))))
```



### Recreate Figure 4.5

```{r}

a0 = seq(0.1, 0.9, 0.03)

pi0_p = sapply(a0, function(a0) {
  
  A0 = qnorm(0.5 + c(-1, 1) * a0/2)
  N = length(prostz)
  N_plus_A0 = length(which(prostz >= A0[1] & prostz <= A0[2]))

  pi0 = N_plus_A0/(N * (pnorm(A0[2]) - pnorm(A0[1])))
  
  return(pi0)
  
})

pi0_D = sapply(a0, function(a0) {
  
  A0 = qnorm(0.5 + c(-1, 1) * a0/2)
  N = length(DTIz)
  N_plus_A0 = length(which(DTIz >= A0[1] & DTIz <= A0[2]))

  pi0 = N_plus_A0/(N * (pnorm(A0[2]) - pnorm(A0[1])))
  
  return(pi0)
  
})


plot(a0, pi0_p, type = 'l', main = 'by 0.03', ylab = 'pi0 estimate')
lines(a0, pi0_D, lty = 2)
legend('bottomright', c('prostate', 'DTI'), lty = c(1, 2), bty = 'n')

### finer detail
a0 = seq(0.1, 0.9, 0.001)

pi0_p = sapply(a0, function(a0) {
  
  A0 = qnorm(0.5 + c(-1, 1) * a0/2)
  N = length(prostz)
  N_plus_A0 = length(which(prostz >= A0[1] & prostz <= A0[2]))

  pi0 = N_plus_A0/(N * (pnorm(A0[2]) - pnorm(A0[1])))
  
  return(pi0)
  
})

pi0_D = sapply(a0, function(a0) {
  
  A0 = qnorm(0.5 + c(-1, 1) * a0/2)
  N = length(DTIz)
  N_plus_A0 = length(which(DTIz >= A0[1] & DTIz <= A0[2]))

  pi0 = N_plus_A0/(N * (pnorm(A0[2]) - pnorm(A0[1])))
  
  return(pi0)
  
})


plot(a0, pi0_p, type = 'l', main = 'by 0.001', ylab = 'pi0 estimate')
lines(a0, pi0_D, lty = 2)
legend('bottomright', c('prostate', 'DTI'), lty = c(1, 2), bty = 'n')

```


# Section 5.2

## Lindsey's method to recreate Figure 5.1 and Figure 5.2 (estimation of false discovery rates) using $\widehat{f}$

The example on page 78 for estimating $\widehat{fdr}$ versus $\bar{fdr}$ is indicated by the pink lines

```{r}
N = 6033 
  # N = length(prostz)
N_plus_A0 = length(which(prostz >= A0[1] & prostz <= A0[2]))
# pi0 = N_plus_A0/(N * (pnorm(A0[2]) - pnorm(A0[1])))
  pi0 = 0.932

z = seq(-4.5, 4.5, by=0.01)

K = 90
d = 0.1
J = 7
breaks = seq(-4.5, 4.5, d)
length(breaks)-1 == K

h = hist(prostz, breaks=breaks, plot=F)
(scale = N*d)

# Lindsey's method
Zk = h$breaks
yk = h$counts
xk = h$mids

# fit = glm(yk ~ poly(xk, J, raw=T), family="poisson")
fit = glm(yk ~ splines::ns(xk, df=J), family="poisson")

yhat_xk = predict(fit, newdata = data.frame(xk = xk), type = 'response')
yhat_z = predict(fit, newdata = data.frame(xk = z), type = 'response')

f0_xk = dnorm(xk, 0, 1)
f0_z = dnorm(z, 0, 1)

## Figure 5.1 
plot(h, main='Prostate data', xlab='z values', ylab='counts')
  # scaled estimate pi0*f0(z)
  lines(z, scale*pi0*f0_z, lty=2)
  # estimate for mixture using Lindsey's method
  lines(z, yhat_z, lty=1)
  abline(h = c(4.49, 8), col = 'lightpink', lty=3)
  abline(v = c(3.2, 3.3), col = 'lightpink', lty=3)
  

```

Estimate local fdr with example bin $k = 78$ (on page 78... wonder if that was a coincidence `r emo::ji("thinking")`). Bin $k=78$ corresponds to $z \in [3.2, 3.3)$, and is observed to contain $y_k = 8$ values. 

The smooth estimate of the local fdr is 

\begin{align}
\widehat{fdr}(x_k) & = \frac{\widehat{\pi}_0 f_0(x_k)}{\widehat{f}(x_k)} \hskip 5mm \text{(5.5)} \\
& =  \frac{\widehat{\pi}_0 f_0(x_k)}{\widehat{f}(x_k)} \cdot \frac{N \cdot d}{N \cdot d} \\
& = \frac{N \cdot \hat{\pi}_0 \cdot f_0(x_k) \cdot d}{N \cdot d \cdot \hat{f}(x_k)} \\
& \approx \frac{N \cdot \widehat{\pi}_0 \cdot [F_0(x_k + d/2) - F_0(x_k - d/2)]}{N \cdot d \cdot \widehat{f}(x_k)} \\
& = \frac{e_{0k}}{\widehat{y}_k}
& = \frac{1.14}{4.49} \\
& = 0.254
\end{align}

```{r}

## Example p.78 - bin k = 78
# equation (5.20)
(e0k = N*pi0*dnorm(xk[78])*d)
(yhat_xk[78])
(fdr_hat_78 = e0k/(yhat_xk[78]))

## ways to calculate the null expected count:  
N*pi0*(pnorm(xk[78]+d/2) - pnorm(xk[78]-d/2)); N*pi0*dnorm(xk[78])*d

## Estimate local fdr [and tail area Fdr]
e0k = N*d*pi0*f0_xk
fdr_hat_xk = e0k/yhat_xk
fdr_bar_xk = e0k/yk
# Fdr = pi0*f0/(...)  
  
## Figure 5.2
plot(xk, pmin(fdr_hat_xk, 1), lty=1, type='l')  

## Table 5.1
k = seq(76, 83)
knitr::kable(cbind(k, 'xk'=xk[k], 'e0k'=e0k[k], 'yk'=yk[k], 'fdr_bar'=fdr_bar_xk[k], 'y_hat'=yhat_xk[k], 'fdr_hat'=fdr_hat_xk[k]), row.names = F)
  
```





* Note: by default, the function `locfdr` replaces the polynomial exponent with a natural spline basis
