---
title: "Chapter 5 example"
output:
  html_document:
header_includes:
  - \usepackage{amsmath, mathtools}
  - \usepackage[mathscr]{euscript}
---
   
# <span style="font-family:Georgia; font-size:0.65em;"> Estimate $\pi_0$ using section 4.5 methods  </span>

The `prostz.RData` consists of 6,033 z-values from the prostate cancer data analysis example. 

```{r}
load(file.path("/Users/valeriewelty/Dropbox/Vandy/SEDS Lab Group/lsi", "prostz.RData"))
length(prostz)
```

In this example, we take $\mathscr{A}_0$ to be the central $\alpha_0$ proportion of the $f_0$ distribution as on page 60. Here we use $\alpha_0 = 0.5$ and $f_0$ is $\mathscr{N}(0, 1)$ such that   

\begin{align}
\mathscr{A}_0 & = \left[ \Phi^{-1} (0.5 - \alpha_0/2), \Phi^{-1}(0.5+\alpha_0/2) \right] \\
& = \left[ \Phi^{-1} (0.5 - 0.5/2), \Phi^{-1}(0.5+0.5/2) \right] \\
& = \left[ \Phi^{-1} (0.25), \Phi^{-1}(0.75) \right] \\
& = \left[ -0.6745, 0.6745 \right] 
\end{align}

```{r, results = "markup"}
(a0 = 0.5)
(A0 = qnorm(0.5 + c(-1, 1) * a0/2))
```

We "believe" that $f_1(z)$ is zero for this subset of the sample space, i.e. $f_1(z) = 0$ for $z \in \mathscr{A} = [ -0.6745, 0.6745 ]$. By this assumption, all non-null cases have $z$-values outside of $\mathscr{A}_0$ (called the _zero assumption_). Recall  $N_+(\mathscr{A}_0)$ is the observed number of $z_i$ values in $\mathscr{A}_0$, and here is equal to 2,790. 

```{r}
length(which(prostz >= A0[1] & prostz <= A0[2]))
```
 
Recall that $E[N_+(\mathscr{A}_0)] = \pi_0 \cdot N \cdot F_0(\mathscr{A}_0)$ (4.45), suggesting the estimator 

\begin{align}
\hat{\pi}_0 & = N_+(\mathscr{A}_0)/(N \cdot F_0(\mathscr{A}_0)) \\
& = \frac{2,790}{6,033 \cdot F_0(\mathscr{A}_0)}
\end{align}

```{r}
N = length(prostz)
N_plus_A0 = length(which(prostz >= A0[1] & prostz <= A0[2]))

(pi0 = N_plus_A0/(N * (pnorm(A0[2]) - pnorm(A0[1]))))
```

For the DTI data set, $N = 15,443$ and $N_+(\mathscr{A}_0) = 7,165$ and therefore $\hat{\pi}_0 = $. 

```{r}
load(file.path("/Users/valeriewelty/Dropbox/Vandy/SEDS Lab Group/lsi", "DTIdata.Rda"))
DTIz = DTIdata[, 'z.value']
  
(N = length(DTIz))
(N_plus_A0 = length(which(DTIz >= A0[1] & DTIz <= A0[2])))
(pi0 = N_plus_A0/(N * (pnorm(A0[2]) - pnorm(A0[1]))))
```



Recreate Figure 4.5: 

```{r}

a0 = seq(0.1, 0.9, 0.03)

pi0_p = sapply(a0, function(a0) {
  
  A0 = qnorm(0.5 + c(-1, 1) * a0/2)
  N = length(prostz)
  N_plus_A0 = length(which(prostz >= A0[1] & prostz <= A0[2]))

  pi0 = N_plus_A0/(N * (pnorm(A0[2]) - pnorm(A0[1])))
  
  return(pi0)
  
})

pi0_D = sapply(a0, function(a0) {
  
  A0 = qnorm(0.5 + c(-1, 1) * a0/2)
  N = length(DTIz)
  N_plus_A0 = length(which(DTIz >= A0[1] & DTIz <= A0[2]))

  pi0 = N_plus_A0/(N * (pnorm(A0[2]) - pnorm(A0[1])))
  
  return(pi0)
  
})


plot(a0, pi0_p, type = 'l', main = 'by 0.03', ylab = 'pi0 estimate')
lines(a0, pi0_D, lty = 2)
legend('bottomright', c('prostate', 'DTI'), lty = c(1, 2), bty = 'n')

### finer detail
a0 = seq(0.1, 0.9, 0.001)

pi0_p = sapply(a0, function(a0) {
  
  A0 = qnorm(0.5 + c(-1, 1) * a0/2)
  N = length(prostz)
  N_plus_A0 = length(which(prostz >= A0[1] & prostz <= A0[2]))

  pi0 = N_plus_A0/(N * (pnorm(A0[2]) - pnorm(A0[1])))
  
  return(pi0)
  
})

pi0_D = sapply(a0, function(a0) {
  
  A0 = qnorm(0.5 + c(-1, 1) * a0/2)
  N = length(DTIz)
  N_plus_A0 = length(which(DTIz >= A0[1] & DTIz <= A0[2]))

  pi0 = N_plus_A0/(N * (pnorm(A0[2]) - pnorm(A0[1])))
  
  return(pi0)
  
})


plot(a0, pi0_p, type = 'l', main = 'by 0.001', ylab = 'pi0 estimate')
lines(a0, pi0_D, lty = 2)
legend('bottomright', c('prostate', 'DTI'), lty = c(1, 2), bty = 'n')

```



